---
title: "Wholesale Elec Data Exploration"
author: "Kynan Witters Hicks"
date: "January 20, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message = FALSE}

library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
library(tsibble)
library(forecast)
library(tseries)
library(xts)
library(plyr)
library(zoo)

```


###Wholesale Electricity Data Exploration

####CAISO Data Wrangle - 2018 Data Only

#####5-minute interval data for three hubs: 
TH_NP15
TH_SP15
TH_ZP26

```{r get and format data, message=FALSE}


#5-minute interval data for three CA electricity hubs. Source = http://www.energyonline.com
caiso <- read_csv("caiso_2018_original.csv") 

#convert "Data" data from character to date/time

caiso$date <- as.POSIXct(caiso$date,
                         format="%m/%d/%Y %H:%M")

```

```{r filter date by hub}

caiso_sp15 <- caiso %>% 
  filter(hub == "TH_SP15")

caiso_np15 <- caiso %>%
  filter(hub == "TH_NP15")

caiso_zp26 <- caiso %>% 
  filter(hub == "TH_ZP26")
```

```{r fill caiso data gaps, message=FALSE}

#sp15 hub

hub1 <- rep("TH_SP15", each=8760)

caiso_sp15_nodup <- caiso_sp15[!duplicated(caiso_sp15$date,
                                 by = c("date","hub")),]

caiso_sp15_clean <- caiso_sp15_nodup %>%
  as_tsibble(key = id(hub), index = date) %>% 
  fill_gaps(price = median(price)) %>% 
  index_by(datetime = floor_date(date, "hour")) %>% 
  summarise(price = mean(price))

caiso_sp15_clean2 <- data.frame(caiso_sp15_clean, hub1) %>% 
  rename(hub = hub1)
  
#np15 hub

hub2 <- rep("TH_NP15", each=8760)

caiso_np15_nodup <- caiso_np15[!duplicated(caiso_np15$date,
                                 by = c("date","hub")),]

caiso_np15_clean <- caiso_np15_nodup %>%
  as_tsibble(key = id(hub), index = date) %>% 
  fill_gaps(price = median(price)) %>% 
  index_by(datetime = floor_date(date, "hour")) %>% 
  summarise(price = mean(price))

caiso_np15_clean2 <- data.frame(caiso_np15_clean, hub2) %>% 
  rename(hub = hub2)

#zp26 hub

hub3 <- rep("TH_ZP26", each=8760)

caiso_zp26_nodup <- caiso_zp26[!duplicated(caiso_zp26$date,
                                 by = c("date","hub")),]

caiso_zp26_clean <- caiso_zp26_nodup %>%
  as_tsibble(key = id(hub), index = date) %>% 
  fill_gaps(price = median(price)) %>% 
  index_by(datetime = floor_date(date, "hour")) %>% 
  summarise(price = mean(price))

caiso_zp26_clean2 <- data.frame(caiso_zp26_clean, hub3) %>% 
  rename(hub = hub3)

#caiso_whole

caiso_nm <- rep("caiso", each = 26280)

caiso_bind <- rbind(caiso_np15_clean2, caiso_sp15_clean2, caiso_zp26_clean2)
caiso_clean <- data.frame(caiso_bind, caiso_nm) %>% 
  rename(iso = caiso_nm)

```

```{r export caiso hourly data}

write.csv(caiso_clean, "caiso_2018_hourly_byhub.csv")

```

```{r combine hubs, message = FALSE}

caiso_single <- read_csv("caiso_2018_hourly_byhub.csv") %>% 
  dplyr::group_by(datetime) %>% 
  dplyr::summarise(
    mean(price)
  ) %>% 
  dplyr::rename(price = "mean(price)")

write.csv(caiso_single, "caiso_2018_hourly_avghubs.csv")

```

####CAISO Data Wrangle - 2015 - 2018 Data

#####5-minute intervale data 

```{r get and format data, message=FALSE}

#Average 5-minute CA wholesale electricity price. Source = http://www.energyonline.com
caiso_4yr <- read_csv("caiso_2015-2018_original.csv")

#convert "Data" data from character to date/time
caiso_4yr$date <- as.POSIXct(caiso_4yr$date,
                         format="%m/%d/%Y %H:%M")

```

```{r fill caiso data gaps, message=FALSE}

#Remove NA Values
caiso_na <- which(is.na(caiso_4yr$date), arr.ind=TRUE)
caiso_nna <- caiso_4yr[-c(caiso_na),]

caiso_hr <- caiso_nna %>% 
  dplyr::group_by(Date = floor_date(date, "1 hour")) %>% 
  dplyr::summarise(
    mean(price)
  ) %>% 
  dplyr::rename(price = "mean(price)")

caiso_day <- caiso_nna %>% 
  dplyr::group_by(Date = floor_date(date, "1 day")) %>% 
  dplyr::summarise(
    mean(price)
  ) %>% 
  dplyr::rename(price = "mean(price)")

write.csv(caiso_day, "caiso_2015-2018_daily.csv")


```

####Energy Data Wrangle

#####Hourly generation data for 5 renewable energy projects (3 solar and 2 wind) for a year

```{r energy_data}

#Renewable production data. Source = https://pvwatts.nrel.gov/pvwatts.php

renew_prod <- read.csv("renewable_gen_info.csv")

#Renewable project meta data. Data simulated. Not real projects

renew_met <- read.csv("renewable_metadata.csv")

#PPA Prices. Source = https://leveltenenergy.com/blog/ppa-price-index/q3-ppa-price-index/
ppa_price <- seq(19,35, by = 1)

```

```{r reformat energy data}

#reformat so in tidy format

Whitewater <- renew_prod %>% 
  dplyr::select(Whitewater, Month, Day, Hour) %>% 
  dplyr::mutate(project_name = "Whitewater") %>% 
  dplyr::rename(production = Whitewater)

Sonoma <- renew_prod %>% 
  dplyr::select(Sonoma, Month, Day, Hour) %>% 
  dplyr::mutate(project_name = "Sonoma")%>% 
  dplyr::rename(production = Sonoma)

Lompoc <- renew_prod %>% 
  dplyr::select(Lompoc, Month, Day, Hour) %>% 
  dplyr::mutate(project_name = "Lompoc")%>% 
  dplyr::rename(production = Lompoc)

Fairfield <- renew_prod %>% 
  dplyr::select(Fairfield, Month, Day, Hour) %>% 
  dplyr::mutate(project_name = "Fairfield")%>% 
  dplyr::rename(production = Fairfield)

Rosamond <- renew_prod %>% 
  dplyr::select(Rosamond, Month, Day, Hour) %>% 
  dplyr::mutate(project_name = "Rosamond")%>% 
  dplyr::rename(production = Rosamond)

renew_prod_final <- rbind(Whitewater, Sonoma, Lompoc, Fairfield, Rosamond)

write.csv(renew_prod_final, "renew_prod_final.csv")

```

####Explore Timeseries Data 

```{r decompose daily data 2015 - 2018, message=FALSE}

caiso_day <- read_csv("caiso_2015-2018_daily.csv")

caiso_ts_day <- ts(caiso_day$price, frequency = 7, start = c(2015,1))

caiso_dc_day <- decompose(caiso_ts_day)
plot(caiso_dc_day)

```

```{r forecast daily data 2015 - 2018, message=FALSE}

caiso_day_pdq <- auto.arima(caiso_day$price)
caiso_day_pdq

#pdq: (1,1,1)

caiso_day_arima <- arima(caiso_ts_day, order = c(1,1,1))
caiso_day_arima

forecast_caiso_day <- forecast(caiso_day_arima, h = 365) #365 days = 1 year
plot(forecast_caiso_day)

caiso_day_forecast_mean <- forecast_caiso_day$mean
caiso_day_forecast_lower <- forecast_caiso_day$lower
caiso_day_forecast_upper <- forecast_caiso_day$upper

date_seq <- seq(ymd('2019-01-01'),ymd('2019-12-31'), by = 'days')

caiso_forecast_df <- data.frame(date_seq,
                                caiso_day_forecast_mean, 
                                caiso_day_forecast_lower, 
                                caiso_day_forecast_upper)

caiso_forecast_df_ed <- caiso_forecast_df %>% 
  dplyr::select(date_seq, caiso_day_forecast_mean, X95., X95..1)

colnames(caiso_forecast_df_ed) <- c("Date", "price", "lower", "upper")

caiso_day_complete2 <- rbind.fill(caiso_day, caiso_forecast_df_ed)

caiso_day_complete2$Date <- as.Date(caiso_day_complete$Date,
                                  format="%m/%d/%Y")

```

```{r explore hourly data in a day}

caiso_day <- read_csv("caiso_2018_hourly_avghubs.csv") %>% 
  dplyr::mutate(hour = (hour(datetime))) %>% 
  dplyr::group_by(hour) %>% 
    dplyr::summarise(
      mean(price)
    )

plot(caiso_day)

```

```{r explore daily data}

caiso_daily <- read_csv("caiso_2018_hourly_avghubs.csv")

caiso_daily$datetime <- as.POSIXct(caiso_daily$datetime,
                         format="%m/%d/%Y %H:%M")

caiso_daily_ed <- caiso_daily %>%
  dplyr::mutate(key = "caiso") %>% 
  as_tsibble(key = id(key), index = datetime) %>% 
  index_by(date = floor_date(datetime, "hour")) %>% 
  summarise(price = mean(price))

plot(caiso_daily)

```

####Revenue Model

$Revenue = Production (kWh) \times Wholesale Price (\$/kWh)$
$Cost = Production (kWh) \times PPA Price (\$kWh)$

####Attempt to Create geom_ribbons for tab 3: Project Revenue
```{r}
#Create values for geom_ribbon 

#Ribbon #1. Seperate data for ribbon only above ppa price and below wholesale price when energy project is generating. Then set max and min

positive_rev <- ifelse(gen_day1$production > 0, 
                       ifelse(wholesale_day1$price > 25,
                              wholesale_day1$price - 25,
                              ""), "")

positive_rev_df <- data.frame(positive_rev, renew_date_string) %>% 
  dplyr::rename(datetime = renew_date_string)

negative_rev <- ifelse(gen_day1$production > 0,
                       ifelse(wholesale_day1$price < 25,
                              25 - wholesale_day1$price,
                              ""), "")

negative_rev$max <- 0
negative_rev$min <- min(negative_rev)

##ggplot 

#Create normalizer for second y-axis (production (MWh))
normalizer <- 1.5

ggplot(wholesale_day1) +
  geom_line(aes(x = datetime, 
                y = price, 
                color = "line1"), 
            size = 1,
            show.legend = TRUE) +
  geom_line(data = gen_day1, 
            aes(x = datetime,
                y = production*normalizer,
                color = "line2"),
            size = 1,
            show.legend = TRUE) +
  geom_hline(aes(yintercept = 25,
                 color = "line3"),
             size = 1,
             show.legend = TRUE) +
  geom_ribbon(data = positive_rev_df, 
              aes(x = datetime, ymin=25, ymax=as.numeric(positive_rev))) +
  scale_colour_manual(name = "Key:",
                      values = c("line1" = "deeppink3",
                                 "line2" = "goldenrod",
                                 "line3" = "dodgerblue3"),
                      labels = c("Wholesale Price",
                                 "Energy Production",
                                 "PPA Price")) +
  scale_y_continuous("Price (USD)", 
                     sec.axis = sec_axis(~./normalizer, 
                                         name = "Production (MWh)",
                                         breaks = c(0,15,30,45,60)),
                     breaks = c(0,20,40,60,80),
                     limits = c(0,80)) +
  scale_x_datetime(date_breaks = "1 hour", 
                   date_labels = "%H",
                   expand = c(0,0)) +
  labs(title = "How Revenue from a vPPA is Generated (An Example Over One Day)",
       x = "Hour") +
  theme_set(theme_bw()) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5, 
                                  size = 18,
                                  margin = margin(0,0,20,0)),
        axis.text = element_text(size = 16),
        axis.title.x = element_text(size = 14,
                                    margin = margin(10,0,0,0)),
        axis.title.y = element_text(size = 14,
                                    margin = margin(0,10,0,10)),
        axis.title.y.right = element_text(size = 14,
                                          margin = margin(0,10,0,10)),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "bottom",
        legend.direction = "horizontal")


```

